---
title: 'Analysing the HIV pandemic, Part 4: Logistic regression'
author: Andrie de Vries and Armand Bester 
date: ''
slug: pipeline-for-analysing-hiv-part-4
categories:
  - R Language
  - R in Medicine
tags:
  - Medicine
  - R/Medicine
  - HIV/AIDS
summary: "This is part 4 of a four-part series about the HIV epidemic in Africa. In a recent publication in PLoS ONE, the authors described how they used affordable hardware to create a phylogenetic pipeline, tailored for the HIV drug-resistance testing facility. In this part, we discuss genetic diversity and how this can be used to analyse inter- and intra- patient sequence classification using logistic regression."
---

```{r setup, include=FALSE}
# packages required for this post
local({
  # CRAN
  pkgs <- c(
    "BiocManager",
    "ape",
    "tibble",
    "tidyr",
    "dplyr",
    "knitr",
    "RColorBrewer",
    "diagram"
  )

  not_installed <- setdiff(pkgs, as.data.frame(installed.packages())$Package)
  if (length(not_installed)) {
    install.packages(not_installed)
  }

  # BioConductor
  pkgs <- c(
      "Biostrings"
  )
  not_installed <- setdiff(pkgs, as.data.frame(installed.packages())$Package)
  if (length(not_installed)) {
    BiocManager::install(not_installed, ask = FALSE)
  }

})
knitr::opts_chunk$set(echo = TRUE)

static_path <- "/post/2019-05-14-analysis-hiv-pandemic-part-3_files"
static_path <- "."
file_path <- "."

```

*Andrie de Vries is the author of "R for Dummies" and a Solutions Engineer at RStudio*

*Phillip (Armand) Bester is a medical scientist, researcher, and lecturer at the [Division of Virology](https://www.ufs.ac.za/health/departments-and-divisions/virology-home), [University of the Free State](https://www.ufs.ac.za), and [National Health Laboratory Service (NHLS)](http://www.nhls.ac.za/), Bloemfontein, South Africa*




## Introduction


The goal of the project is to find a way to tell whether two genetic samples are from the same person or from two different people.

Samples from the same person can have slightly different genetic sequences, due to mutations and other errors. This is especially problematic in comparing samples of genetic material from retro-active viruses.

To help answer this question, we downloaded data from the Los Alamos HIV database, subtype C, POL CDS.

This document:
* plots the data
* describes a simple machine learning model to differentiate between different samples (intra and inter patient).


## Read the data

The data is in two csv files, for intra and inter patient observations. Each observation is the (dis)similary distance between different samples.

Start by reading the data and combining into a single object.

```{r load, message=FALSE}
library(readr)
library(ggplot2)
library(reshape2)
library(dplyr)
library(magrittr)
```

```{r import, warning=FALSE}
pt_distance <- 
  read_csv("dist_sample_10.csv.zip", col_types = "ccdccf")

head(pt_distance)
```

## Plot the data

Next, plot a histogram of the distance between samples.

```{r histogram}
pt_distance %>% 
  ggplot(aes(x = distance, fill = type)) +
  geom_histogram(binwidth = 0.001) +
  facet_grid(type ~ ., scales = "free_y")
```



## Modeling

To distinguish between the two sample types (intra vs inter) is a binary classification problem.

Logistic regression is a simple algorithm for binary classification.  I train a model with `distance` as independent variable, and `type` the dependent variable.  I train on 100,000 observations purely to reduce computation time.


```{r model}
pt_sample <- pt_distance %>% sample_n(1e5)
model <- glm(type ~ distance, data = pt_sample, family = binomial)
```

Note that sometimes the model throws a warning indicating numerical problems.  This happens because the overlap between intra and inter is very small.  If there is a very sharp dividing line between classes, the logistic regression algorithm has problems to converge.

However, in this case the numerical problems doesn't actually cause a practical problem with model itself.

I show the model summary, in case somebody is interested.  The critical point to observe is that the `distance` variable is highly significant (indicated by the ***).

```{r model_summary}
summary(model)
```

Now I can use the model to compute a prediction for some distances (from 0 to 0.15) and create a plot.

```{r predict}
newdata <-  data.frame(distance = seq(0, 0.05, by = 0.001))
pred <- predict(model, newdata, type = "response")
```

```{r predictionplot}

plot_sample <- 
  pt_sample %>% 
  filter(distance <= 0.05)

plot_inter <- 
  pt_sample %>% 
  filter(distance <= 0.05, type == "Inter")
  
plot_intra <- 
  pt_sample %>% 
  filter(distance <= 0.05, type == "Intra")

ggplot() +
  geom_point(data = plot_inter, aes(x = distance, y = 0), alpha = 0.01, col = "blue") +
  geom_point(data = plot_intra, aes(x = distance, y = 1), alpha = 0.01, col = "red") +
  geom_line(data = newdata, aes(x = distance, y = pred)) +
  geom_text(x = 0.04, y = 0.1, label = "Type = inter") +
  geom_text(x = 0.04, y = 0.9, label = "Type = intra") +
  ggtitle("Model results", subtitle = "Predicted probability that class == 'Intra'")
```

Logistic regression essentially fits an s-curve that indicates the probability.  In this case, for small distances (lower than  ~0.01) the probability of being the same person (i.e. type is intra) is almost 100%.  For distances greater than 0.03 the probability of being type intra is almost zero (i.e. the model predicts type inter).

The model puts the distance threshold at approximately 0.025.


## Conclusion

More work to be done here...
